---
title: "darkmix: example of execution"
date: ""
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example of execution

In this notebook we reproduce the example presented in section 7 of the publication:
Ll. Hurtado-Gil et al., *Darkmix: Mixture Models for Detection and Characterization of Dark Matter Halos*, submitted to ApJ

Used R packages and functions:

* `spatstat` and `spatstat.utils`: `owin`, `pp3`, `unique`, `as.im`, `quadscheme`, `box3`, `variablesinformula` and `volume`
* `plotrix`: `draw.circle`
* `LaplacesDemon`: `Blocks` and `LaplacesDemon`
*	`expint`: `gamma` and `gammainc`
* `misc3d`: `kde3d`
* `Base R`: multiple functions (in particular the function `optim`)

## Step 1 -- Setup

Load the necessary functions and packages.

```{r, message=FALSE}
source("Code/darkmix.R")
library(spatstat)
library(spatstat.utils)
library(expint)
library(plotrix)
library(misc3d)
library(LaplacesDemon)
```

Read the data set to create the point process object and its quadrature: 

```{r}
data <- read.table("Data/datacat.txt")
minx <- 18; maxx <- 43; 
miny <- 171; maxy <- 196; 
minz <- 95; maxz <- 102;
clust <- pp3(x=data[,2],y=data[,3],z=data[,4], box3(xrange=c(minx,maxx), yrange=c(miny,maxy), zrange=c(minz,maxz)))
ntile <- c(50,50,14)
ntile <- c(100,100,28)
quad <- quad.3d(clust, ntile)
```

Declare a Finite Mixture Model of 10 cluster components plus background. Use Einasto profile for the clusters:

```{r}
k <- 10
v <- 5
param2 <-c(k+1,rep(c(einasto.model,v),k), const.model,0)
```

## Step 2 -- Initial guess

Initialize the parameters vector and generate an initial guess of the cluster centers:

```{r}
param <- rep(0,k*(v+1))
cent <- centers(clust, h=0.3, ntile=c(25,25,7), k)
arg <- list(centers=cent[[1]], re=1, n=3, mix=1, back=-2)
param <- mixture.model(param, param2, output=c(), action.initial, arg)
param_df <- param2profile(param)
```

The initial guess can be visualized:

```{r message=FALSE}
arg = list(mar=c(0,0,0,0), clust=clust, col=2, pch=16,border=2, cex=0.5, cexl=2, lwd=2)
out <- mixture.model(param, param2, output = NULL, action.draw, arg)
```

## Step 3 -- Likelihood maximization 

Optimize the parameters via the Nelder-Mead algorithm, and save the result.

```{r}
ocf <- optim(param, model.lik, param2=param2, model=mixture.model, clust=clust, quad=quad, all=FALSE); 
param <- ocf$par
```

For further model refinement we can use the `mask.freeze` functions.

```{r, message=FALSE, results='hide'}
param <- mask.freeze(param, param2, mask.adjust_positions(k,v), clust, quad)
param <- mask.freeze(param, param2, mask.adjust_radii(k,v), clust, quad)
param <- mask.freeze(param, param2, mask.adjust_sersic(k,v), clust, quad)
param <- mask.freeze(param, param2, mask.adjust_mix(k,v), clust, quad)
```

The maximum likelihood, AIC and BIC can be calculated with

```{r}
aic_bic <- AIC.BIC(param, param2, clust, quad)
```

However, the resulting parameters are still far from the best fit.
Repeat the previous steps to obtain better results. 
In the following steps we present the results with the already obtained fit.

Visualize again to see the new model:

```{r, message=FALSE}
arg = list(mar=c(3,3,0,0), clust=clust, col=2, pch=16,border=2, cex=0.5, cexl=2, lwd=2)
out <- mixture.model(param, param2, output = NULL, action.draw,arg)
```

When the model is trusted, we can use it to estimate the number of particles per cluster:

```{r}
arg <- list(clust=clust, quad=quad)
pop <- mixture.model(param,param2,output=c(), action.pop, arg)
pop
```

## Step 4 -- Goodness of fit

The goodness of fit can be evaluated with the function `goodness` (calculation might take several minutes).

```{r}
goodness(param, param2, clust, quad, ntile, residual.type="raw", bandwidth=1)
```

The different kernel density fields can be generated at once. 
By order of appearance: absolute residuals, relative residuals, data density field and model density field.

```{r, message=FALSE}
maps <- c("s", "e", "d", "m") 
plot.mm3d(param, param2, clust, quad, residual.type="raw", bandwidth=1, maps=maps, fast=TRUE, col.point=1, scale="log", proj.var="Z", print=FALSE, print.data=TRUE, w=880, h=880, ntile=ntile) 
```

The membership of each particle to the clusters can be obtained and plotted as follows:

```{r}
classification <- membership(param, param2, mixture.model, clust, threshold=0.3, print=FALSE)
```

Another possibility is to generate new samples reproducing the original data set (if a component has very few particles warnings might appear).

```{r, warning=FALSE}
window <- clust$domain
num_particles <- data.frame(table(classification[[1]]$class))
data_generated <- gen.pattern(param, param2, num_particles, window, scl=4)
```

Visualize the new sample with

```{r}
plot(data_generated)
```

We can visualize as well the empirical density profile centered in a given component together with the full model and the single component estimations. 
For example, for component 4:

```{r}
comp <- 4
profiles <- plot.profile(param, param2, clust, comp, xlim=c(0.1,40), nbin=10, flag.legend=TRUE)
```


